Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-15 aistats.tex
Pass on 6.1

The density of an exponential family for a sample $x$ is
\begin{equation}
	 p_\nat(x) = p(x|\nat) = \exp( \langle \nat, T(x) \rangle - \logpart(\nat)) \; ,
	 \label{eq:def_expfamily}
\end{equation}
where  $\nat$ is called natural (or primal) parameter.
It is fully specified by 1) $T: \cX \rightarrow \real^d$, the sufficient statistic,
and 2) a base measure $\nu$ on $\cX$.
Since the exponential is positive, $p$ has the same support as $\nu$.
The log-partition function $\logpart$ acts as a normalization term, since
\begin{align}
    \logpart(\nat) = \log \int e^{\langle \nat, T(x) \rangle} \nu(dx) \;.
\end{align}
This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding and multivariate normal distributions $\cX=\real^d, \nu$ Lebesgue and $T(X)=(X, X X^\top)$.

@@

{\bf Maximum A Posteriori (MAP).}
Given a dataset $\mD_n =(X_1,\dots,X_n)$, we wish to estimate the maximum of the posterior distribution $p(\nat \cond \mD_n) \propto p(\mD_n|\nat)p(\nat)$.
Plugging in~\eqref{eq:def_expfamily},~\eqref{eq:defBregman} and~\eqref{eq:def_prior} yields
\begin{align}
	p(\nat \cond \mD_n)
    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt^\text{MAP}))
    \label{eq:joint_likelihood}
\end{align}
which reaches its maximum at $\MAPt^\text{MAP}$ such that
\begin{align}
    \nabla \logpart(\MAPt^\text{MAP}) = \MAPm^\text{MAP}
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; ,
    \label{eq:defMAP}
\end{align}
where $T_i=T(X_i)$.
When $n_0=0$ (no samples from the prior), we recover the MLE~\eqref{eq:defMLE}.
We write $\MAPt$ for the MAP, and view the MLE as a special case.

@@

We are now ready to formalize the main problem of this paper. Assume we observe a dataset $\mD_n$ drawn i.i.d. from $p(\cdot \cond\nat^*)$, an exponential family distribution
with parameters $\nat^*$.
We wish to quantify how well the MLE or a MAP approximates the true distribution.

A natural way to quantify this is the Kullback-Leibler divergence (KL) $\KL(p_{\nat^*} || p_\nat)$. % should we define it ?
In the well-specified setting, it corresponds to the log-likelihood sub-optimality~\eqref{eq:suboptimalityKL}.
With exponential families, the KL is also a Bregman divergence:
\alignn{
	\KL(p_{\nat^*} || p_\nat)
	 = \bregman(\nat ; \nat^*)
	 = \bregmanconj(\m^* ; \m) \; .
}
The second equality is a general property of Bregman divergences and convex conjugates. % (Be careful about the order of arguments.)
How does this quantity behave when $\hat \nat$ is the MLE or MAP?
 Or in the words of statistical decision theory, what is the \emph{frequentist risk} of these estimators when the loss is the KL divergence?
This is our first problem.

\begin{problem}[Upper-bounding MAP and MLE]
Upper bound the following quantities:
\begin{align}
	\label{eq:bregmanMLE}
	\text{MLE: } \quad &\expect[\mD_n]{\bregmanconj \left (\E_{\nat^*} [T] ;  \inv{n}  \smallsum_i T_i \right )}, \\
	\label{eq:bregmanMAP}
	\text{MAP: } \quad &\expect[\mD_n]{\bregmanconj \left (\E_{\nat^*} [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )},
\end{align}
where the expectation is on the data $\mD_n = (T_1, \dots, T_n)$.
\end{problem}

@@

\paragraph{Optimization.}
With exponential families, MAP can be seen as \emph{an instantiation of stochastic mirror descent (SMD)}, reviewed in App.~\ref{app:SMD}.
More precisely, let us re-write~\eqref{eq:defMAP} as
\alignn{
\m_n = \m_{n-1}- \lr_n (\m_{n-1} - T_n)
\label{eq:mean_update}
}
where $\lr_n := \inv{n_0 + n}$.
Now define stochastic functions $f_X(\nat) = -\log p(X \cond \nat)$ such that $\E[f_X] = f$.
If we further introduce stochastic gradients $g_n(\nat) := \nabla\logpart(\nat) - T_n = \nabla f_{X_n}(\nat)$, then~\eqref{eq:mean_update} becomes
\alignn{
	\nabla\conj(\hat \nat_{n})
	= \nabla\conj(\hat \nat_{n-1}) - \lr_n g_n(\hat \nat_{n-1}),
}
which is the update formula for SMD on $f$ with mirror map $\nabla\logpart$
and step-size schedule $\lr_n$, initialized at $\nat_0$.
In this view, MLE forgets its (arbitrary) initialization after the first step with step-size 1.
The observation MAP $\in$ SMD brings us to our second problem.
\begin{problem}[Convergence rate for SMD]
Find a convergence rate for stochastic mirror descent that applies to conjugate MAP of exponential families such as Gaussians $\cN(\meanp, \sigma^2)$.
\end{problem}

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=.4\textwidth]{figs/gaussians/new_linear_n0=1.pdf}
	\caption{
	Bias-Variance Decomposition for a Gaussian $\cN(m, \sigma^2)$ with $\meanp^*=(0, 1), \meanp_0 = (1,2)$ and $n_0=1$.
	} 
	\label{fig:gaussian_decomposition}
\end{figure}

@@

As we saw in the technical background, MAP can be interpreted as stochastic mirror descent (SMD).
This means that \textbf{1)} we may obtain a convergence rate for MAP from an optimization analysis, and \textbf{2)} any insights gained from the MAP may inform further design and analysis of SMD.
We first review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating three analyses of SMD with the MAP.
\rlp{mention that examples have so far been lacking for relative SMD.}

@@

Recently, these assumptions have been relaxed to the $\stgcvx$-strong convexity and $\smooth$-smoothness of $f$
\emph{relative} to a reference function $\logpart$, defined as
\aligns{
	\stgcvx \cB_{A}(x, y)
	\leq
	\cB_f(x,y)
	\leq
	\smooth \cB_A(x,y) \; .
}
When $\logpart = \norm{\cdot}^2$, we recover standard (Euclidean) smoothness, and gradient descent.
Those conditions ensure the linear convergence of MD with mirror map $\nabla A$,
extending the theory to non-smooth functions $f$ and non-strongly convex potential $\logpart$
\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}.

For exponential families, the log-likelihood perfectly fit into this framework, as
\aligns{
	f(\theta) = A(\theta) - \expect{\lin{T(X), \theta}}
}
is both $1$-smooth and $1$-strongly convex relative to $A$.
So we are left with finding an applicable convergence rate for SMD under relative assumptions.
As we investigate in the next section, several pieces of work have addressed this question, under various assumptions on the stochasticity \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}.

\subsection{Bounding the Randomness}
\TODO{compress this section}

To analyze stochastic algorithms, one also need to quantify the randomness.
Many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}.
We now review attempts at analyzing SMD under relative smoothness assumption, in the light of  the MAP example.
See \cref{tbl:assumptions} for a summary.

\textbf{Caveat:} these rates apply to the Bregman in the other direction, e.g. $\bregman(\nat^*, \MAPt)$, and sometimes not on the last iterate, but with tail averaging.

\begin{table}[t]
\begingroup
%\renewcommand*{\arraystretch}{1.25}%
\newcommand*{\greencmark}{\textcolor{Green}{\cmark}}
\newcommand*{\redxmark}{\textcolor{Red}{\xmark}}
\centering
\caption{Summary of results for SMD
under relative smoothness and relative strong convexity assumptions.
Barrier tells whether the assumption can hold with barrier objectives.
$\lr_n \sim \inv{n}$ says whether the convergence rate hold with this step-size scheduling, and the last column tells whether the rate applies to the last iterate or to some average iterate.
}
\begin{tabular}{lccc}
%{p{.15\textwidth}m{.05\textwidth}m{.08\textwidth}m{.1\textwidth}}
\toprule
Boundedness & Barriers &  $\lr_n \sim \inv{n}$ & Last iterate \\
\midrule
%Strongly convex and bounded variance\newline
%\citep[e.g.,][]{??}
%&
%$
%\cB_A(\theta,\theta') \geq \frac{1}{2}\norm{\theta-\theta'}^2,
%\quad
%\expect[x]{\norm{\m - x}}^2\leq\sigma^2
%$
%\\
Var. on $\Theta$~\eqref{eq:hanzely} % \newline \citep{hanzely2018fastest}
& \redxmark & \greencmark  & \redxmark
\\
Var. at $\theta_*$~\eqref{eq:dragomir} %\newline \citep{dragomir2021fast}
& \redxmark & \greencmark  & \greencmark
\\
Opt. gap~\eqref{eq:dorazio} %\newline \citep{dorazio2021stochastic}
& \greencmark & \redxmark & \greencmark
\\
\bottomrule
\end{tabular}
\label{tbl:assumptions}
\endgroup
\end{table}

@@

\section{REVIEW OF SMD}
\label{app:SMD}

>>>>

The density of an exponential family for a sample $x$ is
\begin{equation}
	 p_\nat(x) = p(x|\nat) = \exp( \langle \nat, T(x) \rangle - \logpart(\nat)) \; ,
	 \label{eq:def_expfamily}
\end{equation}
where  $\nat$ is called natural (or primal) parameter.
It is fully specified by 1) $T: \cX \rightarrow \real^d$, the sufficient statistic,
and 2) a base measure $\nu$ on $\cX$.
Since the exponential is positive, $p$ has the same support as $\nu$.
The log-partition function $\logpart$ acts as a normalization term, since
\begin{align}
    \logpart(\nat) = \log \int e^{\langle \nat, T(x) \rangle} \nu(dx) \;.
\end{align}
This simple model encompasses both categorical distributions : $\cX = \{1, \dots, k\}$, $\nu$ uniform and $T(X)$  the one-hot encoding, and multivariate normal distributions $\cX=\real, \nu$ Lebesgues and $T(X)=(X, X^2)$.

@@

{\bf Maximum A Posteriori (MAP)}
Given a dataset $\mD_n =(X_1,\dots,X_n)$, we wish to estimate the maximum of the posterior distribution $p(\nat \cond \mD_n) \propto p(\mD_n|\nat)p(\nat)$.
Plugging in~\eqref{eq:def_expfamily},~\eqref{eq:defBregman} and~\eqref{eq:def_prior} yields
\begin{align}
	p(\nat \cond \mD_n)
    \propto \exp(- (n_0+n) \bregman(\nat; \MAPt^\text{MAP}))
    \label{eq:joint_likelihood}
\end{align}
which reaches its maximum at $\MAPt^\text{MAP}$ such that
\begin{align}
    \nabla \logpart(\MAPt^\text{MAP}) = \MAPm^\text{MAP}
    = \frac{n_0 \meanp_0 + \sum_{i=1}^n T_i}{n_0+n} \; ,
    \label{eq:defMAP}
\end{align}
where $T_i=T(X_i)$.
When $n_0=0$ (no samples from the prior), we recover the MLE~\eqref{eq:defMLE}.
We write $\MAPt$ for the MAP, and view the MLE as a special case.

@@

We are now ready to formalize the main problem of this paper. Assume we observe a dataset $\mD$ drawn i.i.d. from $p(\cdot \cond\nat^*)$, an exponential family distribution
with parameters $\nat^*$.
We wish to quantify how well the MLE or a MAP models the true distribution.

A natural way to quantify this is the Kullback-Leibler divergence (KL) $\KL(p_{\nat^*} || p_\nat)$. % should we define it ?
In the well-specified setting it corresponds to the log-likelihood sub-optimality~\eqref{eq:suboptimalityKL}.
With exponential families, the KL is also a Bregman divergence:
\alignn{
	\KL(p_{\nat^*} || p_\nat)
	 = \bregman(\nat ; \nat^*)
	 = \bregmanconj(\m^* ; \m) \; .
}
The second equality is a general property of Bregman divergences and convex conjugates. % (Be careful about the order of arguments.)
How does this quantity behave when $\nat$ is the MLE or MAP?
 Or in the words of statistical decision theory, what is the \emph{frequentist risk} of these estimators when the loss is the KL divergence?
This is our first problem.

\begin{problem}[Upper-bounding MAP and MLE]
Upper bound the following quantities:
\begin{align}
	\label{eq:bregmanMLE}
	\text{MLE: } \quad &\expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )}, \\
	\label{eq:bregmanMAP}
	\text{MAP: } \quad &\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )},
\end{align}
where the expectation is on the data $\mD = (T_1, \dots, T_n)$.
\end{problem}

@@

\paragraph{Optimization}
With exponential families, MAP can be seen as \emph{an instantiation of stochastic mirror descent (SMD).}
More precisely, let us re-write~\eqref{eq:defMAP} as
\alignn{
\m_n = \m_{n-1}- \lr_n (\m_{n-1} - T_n)
\label{eq:mean_update}
}
where $\lr_n := \inv{n_0 + n}$.
Now define stochastic functions $f_X(\nat) = -\log p(X \cond \nat)$ such that $\E[f_X] = f$.
If we further introduce stochastic gradients $g_n(\nat) := \nabla\logpart(\nat) - T_n = \nabla f_{X_n}(\nat)$, then~\eqref{eq:mean_update} becomes
\alignn{
	\nabla\conj(\hat \nat_{n})
	= \nabla\conj(\hat \nat_{n-1}) - \lr_n g_n(\hat \nat_{n-1}),
}
which is the update formula for SMD on $f$ with mirror map $\nabla\logpart$
and step-size schedule $\lr_n$, initialized at $\nat_0$.
In this view, MLE forgets its (arbitrary) initialization after the first step with step-size 1.
The observation MAP$\in$SMD brings us to our second problem.
\begin{problem}[Convergence rate for SMD]
Find a convergence rate for stochastic mirror descent that applies to conjugate MAP of exponential families such as Gaussians $\cN(\meanp, \sigma^2)$.
\end{problem}

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=.4\textwidth]{figs/gaussians/new_linear_n0=1.pdf}
	\caption{
	Bias-Variance Decomposition for a Gaussian $\cN(m, \sigma^2)$ with $\meanp^*=(0, 1), \meanp_0 = (1,2)$ and $n_0=1$. The asymptote is $\frac{1}{n}$.
	} 
	\label{fig:gaussian_decomposition}
\end{figure}

@@

As we saw in \S\ref{sec:problem}, MAP can be interpreted as stochastic mirror descent (SMD).
This means that \textbf{1)} we may obtain a convergence rate for MAP from an optimization analysis, and \textbf{2)} any insights gained from the MAP may inform further design and analysis of SMD.
We first review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating recent analyses of SMD with the MAP.

@@

Recently, these assumptions have been relaxed to the $\stgcvx$-strong convexity and $\smooth$-smoothness of $f$
\emph{relative} to a reference function $\logpart$, defined as
\aligns{
	\stgcvx \cB_{A}(x, y)
	\leq
	\cB_f(x,y)
	\leq
	\smooth \cB_A(x,y) \; .
}
When $\logpart = \norm{\cdot}^2$, we recover standard (Euclidean) smoothness, and gradient descent.
These relative conditions ensure the linear convergence of MD with mirror map $\nabla A$
\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively},
even when $f$ is not smooth and $\logpart$ is not strongly convex.

For exponential families, MAP perfectly fits into this framework, as
\aligns{
	f(\theta) = A(\theta) - \expect{\lin{T(X), \theta}}
}
is both $1$-smooth and $1$-strongly convex relative to $A$.
Our goal is then to find an applicable convergence rate for SMD under relative smoothness.
Several pieces of work have addressed this question, under various assumptions on the stochasticity \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}, 
but they have so far been lacking concrete examples.
In the next section, we challenge their hypothesis thanks with the MAP.

\subsection{Bounding the Randomness}

To analyze stochastic algorithms, one also need to quantify the randomness.
While many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}, only a few have been adapted to SMD with relative smoothness.
We review these in the light of  the MAP and provide a summary in \cref{tbl:assumptions}.

\begin{table}[t]
\newcommand*{\greencmark}{\textcolor{Green}{\cmark}}
\newcommand*{\redxmark}{\textcolor{Red}{\xmark}}
\caption{Summary of results for SMD
under relative smoothness and relative strong convexity assumptions.
Each row correspond to one analysis, and each columns answers one question.
($-\log$) does the bound hold for the Gaussian variance example (\S\ref{ssec:gaussian-variance})?
($\lr_n \sim \inv{n}$) does it converge with a $O(\inv{n})$ step-size?
($f$) is the bound in function value, or in reverse Bregman $\bregman(\nat^*,\hat\nat_n)$? 
 ($\hat\nat_n$) is it for the last iterate or an average ?
None of these analysis check all the boxes needed to address~\eqref{problem}.
}
\begin{center}
\begin{tabular}{lcccc}
%{p{.15\textwidth}m{.05\textwidth}m{.08\textwidth}m{.1\textwidth}}
\toprule
Boundedness & $-\log$ &  $\lr_n \sim \inv{n}$ & $f$ & $\hat\nat_n$ \\
\midrule
%Strongly convex and bounded variance\newline
%\citep[e.g.,][]{??}
%&
%$
%\cB_A(\theta,\theta') \geq \frac{1}{2}\norm{\theta-\theta'}^2,
%\quad
%\expect[x]{\norm{\m - x}}^2\leq\sigma^2
%$
%\\
Var. on $\Theta$~\eqref{eq:hanzely} % \newline \citep{hanzely2018fastest}
& \redxmark & \greencmark & \greencmark  & \redxmark
\\
Var. at $\theta_*$~\eqref{eq:dragomir} %\newline \citep{dragomir2021fast}
& \redxmark & \greencmark & \redxmark  & \greencmark
\\
Opt. gap~\eqref{eq:dorazio} %\newline \citep{dorazio2021stochastic}
& \greencmark & \redxmark & \redxmark & \greencmark
\\
\bottomrule
\end{tabular}
\end{center}
\label{tbl:assumptions}
\end{table}

@@
